\normalfont{A bird's-eye view on the habitability of exoplanets via statistical learning techniques}

### 2020-10-28 Manuscript/Manuscript.tex
PUB

Only rocky planets are considered habitable \footnote{In principle also moons have to be considered, however for sake of simplicity here we take into account only planets}, so gaseous ones such as Jupiter or Saturn have to be considered not habitable (see Fig. \ref{exoplanet_populations} and \ref{PT_Confirmed}); second, as requested by all form of life in the Earth, an habitable planet requires that liquid water can be present \cite{seager2013exoplanet,mckay2014requirements,rothschild2001life} \footnote{Astrobiologist scholars also proposed that other form of life can come up without liquid water (see for instance \cite{rahm2016polymorphism}) however, for simplicity, here only planets that allows liquid water are considered habitable}. This feature is given, at first glance, by the radiation intensity $I$ that is bounded to the power radiation ($P$) produced by the star that host the planet and the respective distance $r$. Indeed by approximating a star as a point or a perfect sphere, a radiation with a spherical symmetry is produced; therefore the integral that gives the power produced $P$: 

@@

P=\int \textbf{I}\cdot \textit{d\textbf{A}}

@@

P=\lvert I \rvert \cdot A_{surf}=4\pi r^{2}

@@

\lvert I \rvert=\dfrac{P}{4\pi r^{2}}

@@

Such relation establishes an interval for the distance $r$ where the stellar radiation is not too hot forcing water is the gas phase, and not to low so that only the solid state phase is allowed. In the literature this interval is usually called habitable-zone \cite{kasting1993habitable}. However this represents an over-simplified model since no atmospheric effect is taken in to account: indeed the surface temperature is also influenced by the greenhouse gases that allows to the host radiation to enter inside the planet but limits its escape \cite{seager2013exoplanet,forget2014possible}. Basically, as explained in figure \ref{Greenhouse-effect-t2} such phenomena is given by the greenhouse molecules, such as CO$_{2}$, vapour, CH$_{4}$, that are excited by the infra-red radiation produced by the planet. Since these remit the energy gained in all directions, the overall heat that is lost by the planet through infra-red radiation is reduced \cite{pierrehumbert2011infrared}, thus the surface temperature is increased. It is worth nothing that, in principle, the greenhouse effect does not reduce the habitability: on the contrary it allows to the planet to keep part of the heath received from the host star. On the other side, if the concentration of the greenhouse gases is too hight, and therefore the surface temperature abruptly increases, the planetary homoeostasis\cite{lovelock1974atmospheric,lovelock1982life,caldeira1992life} breaks and the water that is present on the planet escapes from it; this was probably the fate of Venus that lost its oceans \cite{way2016venus,luger2015extreme,seager2013exoplanet}. The conventional habitable zone can be extended if, for instance, we take in to account the fact that more massive planets are able to bound the H$_{2}$ (a greenhouse gas see Fig \ref{Planets_habitability_Seager}); therefore also the mass of a planet, in principle should be considered as a factor for its habitability. Scholars \cite{tackley2012habitable} also focused on the tectonic phenomena that allows the recycling of atmospheric gases and produces a negative feedback for surface temperature \cite{walker1981negative}: this feature, at a first glance, is largely influenced by the presence of water \cite{korenaga2010likelihood,o2007geological} and weakly by the planet radius \cite{o2007geological} and mass \cite{korenaga2010likelihood}\footnote{The features reported here, such as the stellar luminosity, represent what we can observe now from the planets and stars: since the speed of light is finite ($3\cdot10^{8}$ m/s ) and many stellar systems are vary far away (for instance 582 light years for Kepler-186) such features describe the planets and stars as these were also centuries ago: therefore the habitability refers to the moment when the light signal was broadcast and not to the present situation.} .

>>>>

Only rocky planets are considered habitable \footnote{In principle also moons have to be considered, however for sake of simplicity here we take into account only planets}, so gaseous ones such as Jupiter or Saturn have to be considered not habitable (see Fig. \ref{exoplanet_populations} and \ref{PT_Confirmed}); second, as requested by all form of life in the Earth, an habitable planet requires that liquid water can be present \cite{seager2013exoplanet,mckay2014requirements,rothschild2001life} \footnote{Astrobiologist scholars also proposed that other form of life can come up without liquid water (see for instance \cite{rahm2016polymorphism}) however, for simplicity, here only planets that allows liquid water are considered habitable}. This feature is given, at first glance, by the radiation intensity $I$ that is bounded to the luminosity ($L$) produced by the star that host the planet and the respective distance $R$. Indeed by approximating a star as a point or a perfect sphere, a radiation with a spherical symmetry is produced; therefore the integral that gives the luminosity $L$: 

@@

L=\int \textbf{I}\cdot \textit{d\textbf{A}}

@@

L= I \cdot A_{surf}=4\pi R^{2}

@@

 I =\dfrac{L}{4\pi R^{2}}
\label{Flux}

@@

The latter is called stellar flux. On this basis one can derive the planetary equilibrium temperature starting from the balance equation that equals the flux absorbed by the planet with the emitted one \cite{catling2017atmospheric}: 
F_{abs}=F_{emit}
however since part of incident flux is reflected according to the planet Albedo A, the previous equation can be rewritten in the following way \cite{catling2017atmospheric}:
(1-A_{B})F_{solar}=F_{emit}
furthermore, since only a half hemisphere of a planet can absorb the flux emitted by the half of the star that emits it, we have \cite{catling2017atmospheric}:
F_{solar}=\dfrac{I_{0}}{4}
where I$_{0}$ is the incident radiation on the planet (in this case a scalar). Then plugging the Stefan–Boltzmann law for the radiation of a black-body we obtain \cite{catling2017atmospheric}: 
\left(1-A_{B}\right)\dfrac{I_{0}}{4}=\sigma T^{4}_{eq}
Where $\sigma$ is a constant. Thus the planetary equilibrium temperature will be \cite{catling2017atmospheric}:
T_{eq}=\left(\dfrac{I_{0}\left(1-A_{B}\right)}{4\sigma}\right)
Finally if we plug in the following relation the Eq. \ref{Flux} we obtain \cite{catling2017atmospheric}:
T_{eq}=\left(\dfrac{L\left(1-A_{B}\right)}{16\sigma\pi a^{2}}\right)^{\dfrac{1}{4}}
that can be cast as function of stellar temperature  T$_{star}$ \cite{catling2017atmospheric}:
T_{eq}=T_{star}\sqrt{R/2a}\left(1-A\right)^{0.25}
Such relation establishes an interval for the distance $R$ where the stellar radiation, here given in function of $P$ or $T_{eq}$, is not too hot forcing water is the gas phase, and not to low so that only the solid state phase is allowed. In the literature this interval is usually called habitable-zone. \cite{kasting1993habitable}. However this represents an over-simplified model since no atmospheric effect (except for the albedo) is taken in to account: indeed the surface temperature is also influenced by the greenhouse gases that allows to the host radiation to enter inside the planet but limits its escape \cite{seager2013exoplanet,forget2014possible}. Basically, as explained in figure \ref{Greenhouse-effect-t2} such phenomena is given by the greenhouse molecules, such as CO$_{2}$, vapour, CH$_{4}$, that are excited by the infra-red radiation produced by the planet. Since these remit the energy gained in all directions, the overall heat that is lost by the planet through infra-red radiation is reduced \cite{pierrehumbert2011infrared}, thus the surface temperature is increased. It is worth nothing that, in principle, the greenhouse effect does not reduce the habitability: on the contrary it allows to the planet to keep part of the heath received from the host star. On the other side, if the concentration of the greenhouse gases is too hight, and therefore the surface temperature abruptly increases, the planetary homoeostasis  \cite{lovelock1974atmospheric,lovelock1982life,caldeira1992life} breaks and the water that is present on the planet escapes from it; this was probably the fate of Venus that lost its oceans \cite{way2016venus,luger2015extreme,seager2013exoplanet}. The conventional habitable zone can be extended if, for instance, we take in to account the fact that more massive planets are able to bound the H$_{2}$ (a greenhouse gas see Fig \ref{Planets_habitability_Seager}); therefore also the mass of a planet, in principle should be considered as a factor for its habitability. Scholars \cite{tackley2012habitable} also focused on the tectonic phenomena that allows the recycling of atmospheric gases and produces a negative feedback for surface temperature \cite{walker1981negative}: this feature, at a first glance, is largely influenced by the presence of water \cite{korenaga2010likelihood,o2007geological} and weakly by the planet radius \cite{o2007geological} and mass \cite{korenaga2010likelihood}\footnote{The features reported here, such as the stellar luminosity, represent what we can observe now from the planets and stars: since the speed of light is finite ($3\cdot10^{8}$ m/s ) and many stellar systems are vary far away (for instance 582 light years for Kepler-186) such features describe the planets and stars as these were also centuries ago: therefore the habitability refers to the moment when the light signal was broadcast and not to the present situation.} .
